---
title: "`glmnet` in R"
author: "Torben"
date: "November 24, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = NA, fig.width = 10)
```

Load `glmnet` to make the functionalities available.
```{r, echo=FALSE}
## install.packages("glmnet") ## Most recent CRAN version: 2.0-10
library(glmnetUtils) ## loads library(glmnet)
library(tidyverse)
library(plotmo) # for plot_glmnet # install.packages("plotmo")
library(ggfortify) # for enhancement of ggplot # install.packages("plotmo")
theme_set(theme_bw(base_size = 20))
```

Ressources on `glmnet`

* The package vignette `vignette("glmnet_beta")` is highly recommendable. 
* Many questions have already been asked _and_ answered at https://stackoverflow.com/questions/tagged/glmnet.

## Example

```{r}
crime <- read_csv("crime.csv", col_types = cols(
  funding = col_integer(),
  hs = col_integer(),
  `not-hs` = col_integer(),
  college = col_integer(),
  college4 = col_integer(),
  `crime rate` = col_integer()
)) %>% rename(rate = `crime rate`)
```

Note: Due to different scales of the variables, the estimated parameters may be different in size simply due to
different units. Hence, The `glmnet` function automatically standardises both the response 
(for `family = "gaussian"`) `y` and the covariates `x`.

```{r}
crime_lasso <- glmnet(rate ~ .,
                      data = crime,
                      alpha = 1) ## alpha = 1: LASSO

plot(crime_lasso)
```

```{r}
plot_glmnet(crime_lasso, xvar = "norm")
```

```{r}
autoplot(crime_lasso)
```


```{r}
crime_ridge <- glmnet(rate ~ .,
                      data = crime,
                      alpha = 0) ## alpha = 0: Ridge Regression
plot(crime_ridge)
plot_glmnet(crime_ridge, xvar = "norm")
```

# What should $\lambda$ be?

```{r}
cv_glmnet_lasso <- cv.glmnet(y = crime_rate, x = crime_matr, alpha = 1)
plot(cv_glmnet_lasso)
```


```{r}
cv_glmnet_ridge <- cv.glmnet(y = crime_rate, x = crime_matr, alpha = 0) ## alpha = 0: Ridge
plot(cv_glmnet_ridge)

```

# Elastic net approach: What should $\alpha$ be?

```{r}
cv_glmnet_alpha <- cva.glmnet(rate ~ ., data = crime)
alphas <- cv_glmnet_alpha$alpha
plot(cv_glmnet_alpha, col = 1:11)
legend("topleft", bty = "n",
       lty = 1,
       col =  topo.colors(length(alphas)), 
       title = expression(alpha),
       legend = alphas)
```

# Other types of regression

There are many different types of regressions one would be interested in:

* Logistic regression (binary outcomes): `family = "binomial"`
* Count regression (Poisson distribution): `family = "poisson"`
* Multiclass (multinomial data): `family = "multinomial"`
* Survival analysis (Cox proportional hazard model): `family = "cox"`
* Multivariate normal (multivariate Gaussian): `family = "mgaussian"`



