---
title: "Tibbles (Chapter 7)"
author: "Torben"
date: "November 22, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = NA)
```

```{r}
library(tidyverse)
```

# Tibbles

Tibbles are the modern alternative to the `data.frame` structure.

Not only does it have prettier print methods
```{r}
mtcars %>% print()
mtcars %>% as_tibble() %>% print()
## mtcars %T>% print() %>% as_tibble() %>% print()
```

It also extends to types of elements in the `tibble` - 
e.g. we could have a `tibble` inside a `tibble` (and most other data structures) 
by the use of the `list` column type.

```{r}
X <- tibble(x = list(1:4),
       y = 1,
       z = "A")
X
X %>% pull(x)
```

The main difference between the `tidyverse` approach and the *older* type of R programming,
is the decreased use of `[]` and `[[]]` for subsetting in dataframes.

```{r}
mtcars[mtcars$cyl==4,c("cyl", "hp")]
## subset(mtcars, cyl == 4, c("cyl", "hp")) also works
mtcars %>% filter(cyl == 4) %>% select(cyl,hp)
```

Note however, `tibbles` do not allow for `rownames`

# Exercises

From chapter 7 solve the following exercises (pp 123 - 124):

1, 2, 3, 5 (and 6)