---
title: "Factors with forcats (Chapter 12)"
author: "Torben"
date: "November 22, 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = NA)
```

```{r}
library(tidyverse)
theme_set(theme_bw())
library(forcats)
```

# Factors

Factors are string-like classes that where introduced in R to ease e.g. string comparisons,
as factors are **categorical variables** only allowed to take on a limited number of defined states.

They are essential in the way we think of statistical modelling in R.

However, they can be a **pain** to deal with for many non-modelling purposes.

The `forcats` package introduces a number of helper functions to ease working with factors.

```{r}
x1 <- c("Dec", "Apr", "Jan", "Mar")
x2 <- c("Dec", "Apr", "Jam", "Mar")
month_levels <- c(
  "Jan", "Feb", "Mar", "Apr", "May", "Jun", 
  "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"
)
```

To tell R there is a non-lexicographic ordring we provide the levels in their natural ordering
```{r}
sort(x1)
sort(factor(x1, levels = month_levels))
```

If a case is not matched by the predefined classes we get a `NA` in the result
```{r}
factor(x2, levels = month_levels)
```

Sometimes the ordering in the datafile should be used as ordering of the levels.
The `fct_inorder` handles this -- functions starting with `fct_` most likely originates from `forcats`
```{r}
x1 %>% factor() %>% fct_inorder()
```

# Example

We look at the `gss_cat` data as in the book. To reverse the levels of the e.g. `race` factor, we 
can use the `fct_rev` function can combine this with a tabulation of race and marital status

```{r}
gss_cat %>%
  mutate(race = fct_rev(race)) %>% 
  count(race, marital)
```

To get this into a more tabular format we can add a spread line in the end
```{r}
gss_cat %>%
  mutate(race = fct_rev(race)) %>% 
  count(race, marital) %>% 
  spread(key = marital, value = n)
```

## Modifying the ordering 

It may be instructive to reorder a factor based on the marginal information of one or two variables.
E.g. some plots convey a stronger message if the trend in data is visualised by reordering the groups.

```{r}
relig_summary <- gss_cat %>%
  group_by(relig) %>%
  summarise(
    age = mean(age, na.rm = TRUE),
    tvhours = mean(tvhours, na.rm = TRUE),
    n = n()
  )

ggplot(relig_summary, aes(tvhours, relig)) + geom_point()
```

We clearly see that ordering the data according to the TV hours is more informative.
For this the `fct_reorder` is perfect -- without it we would need a lot of taylored code to rearrange the `relig` factor.
Below the `relig` factor is ordered such that the levels are in increasing order according to `tvhours`.

```{r}
ggplot(relig_summary, aes(tvhours, fct_reorder(relig, tvhours))) +
  geom_point()
```

The function `fct_relevel` makes it easy to put a specific value in a specific spot in the levels
```{r}
days <- c("Mon", "Tue", "Weds", "Thur", "Fri", "Sat", "Sun") %>%
  factor() %>% fct_inorder()
days
fct_relevel(days, "Sun", after = 0) ## Week starts with Sunday
```

## Modifying the levels

Changing the names of the levels is typically necessary for pretty plots and tablular output.

The `fct_recode` and `fct_collapse` are friends in need for this recurring challenge.

To recode/rename we combine `mutate` and `fct_recode`, where the syntax is
`"new level" = "old level"` in `fct_recode` with the first argument being the factor to modify.

```{r}
gss_cat %>%
  mutate(partyid = fct_recode(partyid,
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"
  )) %>%
  count(partyid)
```

Typically we have made a two fine split of the data and wants to collapse these to fewer categories.
Here `fct_collapse` is handy -- it has the same syntax as `fct_recode`.

```{r}
gss_cat %>%
  mutate(partyid = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep = c("Strong republican", "Not str republican"),
    ind = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem = c("Not str democrat", "Strong democrat")
  )) %>%
  count(partyid)
```

The last bit of tweeking is to "lump" together the smallest groups in the data.
Using the `fct_lump` we can collapse the variable into the `n` largest groups and the group `other_level` 
(by default called `"Other"`)

```{r}
gss_cat %>%
  mutate(relig = fct_lump(relig, n = 10, other_level = "Other")) %>%
  count(relig, sort = TRUE) %>%
  print(n = Inf)
```

