---
title: "Data Import with readr  (Chapter 8)"
author: "Torben"
date: "November 22, 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, message = FALSE)
```

```{r}
library(tidyverse)
```

# `readr`

The comma separated files are a typical flat spreadsheet file format to encounter (we discuss Excel files below).

The `read_csv` and `read_csv2` functions are replacements for base R's 
`read.csv` and `read.csv2` with almost the same properties. However, there are some important differences.

With the `_` functions strings are __never__ converted to factors. This is default for the `.` counterparts!

```{r, include=FALSE}
## Saves the massey-rating.txt from the readr package as local file
write_csv(read_table(readr_example("massey-rating.txt")), path = "massey_rating.csv")
```

Read in some example data with numeric and string columns
```{r, message=TRUE}
massey_ <- read_csv("massey_rating.csv")
massey. <- read.csv("massey_rating.csv")
```

For `massey.` the strings are converted to factors -- can cause problems down stream in your analysis.
```{r}
class(massey_$Team)
class(massey.$Team)
```

## Column types 

From the output above we got a parsing message informing us about the type of class of each column.
By default the `col_guess()` is used - this function *guesses* the column type by the first `guess_max` rows in the file.

We can change this using the `col_types` argument in `read_csv` by using `cols` and the `col_*` functions:

* `col_character` (`"c"`)
* `col_double` (`"d"`)
* `col_factor`
* `col_guess`  (`"?"`)
* `col_integer` (`"i"`)
* `col_logical` (`"l"`)
* `col_number` (`"n"`)

* `col_date` (`"D"`)
* `col_datetime` (`"T"`)
* `col_time` (`"t"`)

For example, we may insist that `UCC` is of type double by using `col_double()`

```{r}
read_csv("massey_rating.csv", 
         col_types = cols(
           UCC = col_double()
           ))
```

This can be shortened by the shorthand `"d"`
```{r}
read_csv("massey_rating.csv", 
         col_types = cols(
           UCC = col_double(),
           PAY = "d"
           ))
```

The special `col_skip` (`"_"` or `"-"`) discards the column on import
```{r}
read_csv("massey_rating.csv", 
         col_types = cols(
           DII = "_", ## Skip
           ENG = "-", ## Skip
           ACU = "c",
           Rank = "i",
           Team = col_factor(NULL),
           Conf = col_character()
           )
         )
```

## Only specific 

If it is only a small part of the input columns we would like to read, then the `cols_only` is efficient
```{r}
read_csv("massey_rating.csv", 
         col_types = cols_only(
           Team = col_factor(NULL),
           Conf = "c"
           )
         )
```

## Column names

If the input file do not have column headers, we can provide these and set their type.

```{r}
read_csv("1,2,3\n4,5,6",
         col_names = c("A", "B", "C"),
         col_types = cols(
           A = "c",
           B = "i",
           C = "d"
         ))
```


Other functions in `readr` for other types of flat files:

* `read_tsv()`: tab separated files
* `read_delim()`: general delimited files
* `read_fwf()`: fixed width files
* `read_table()`: tabular files where columns are separated by white-space.
* `read_log()`: web log files

**`data.table` (a different package with a _very_ different syntax)**

data.table has a function similar to `read_csv()` called `fread`. Compared to `fread`, `readr` functions:

* Are slower (currently ~1.2-2x slower. If you want absolutely the best performance, use `data.table::fread()`.

* Use a slightly more sophisticated parser, recognising both doubled (`""""`) and backslash escapes (`"\""`), and can produce factors and date/times directly.

* Forces you to supply all parameters, where `fread()` saves you work by automatically guessing the delimiter, whether or not the file has a header, and how many lines to skip.

* Are built on a different underlying infrastructure. Readr functions are designed to be quite general, which makes it easier to add support for new rectangular data formats. `fread()` is designed to be as fast as possible.

# `readxl`

Reads data from Excel (`xls` and `xlsx` types) with the possibility to select different sheets and ranges.

# `haven` 

To interact with SPSS (`read_spss` or `read_sav`), stata (`read_stata`) and SAS (`read_sas`) data files

# `rio`

Attempt to harmonise the interface for read and write from/to external files for various file formats.

https://github.com/leeper/rio



```{r, include=FALSE}
## Deletes the massey_rating.csv file again
file.remove("massey_rating.csv")
```
