---
output: html_document
---

# Vectors

## Introduction

* tibbles and packages that work with them
* More programming (own functions), dig deeper into R: vectors

### Prerequisites

Use functions from the __purrr__ package to avoid some inconsistencies in base R.

```{r setup, message = FALSE}
library(tidyverse)
```

## Vector basics

There are two types of vectors:

1. __Atomic__ vectors, of which there are six types:
  __logical__, __integer__, __double__,  __character__, __complex__, and 
  __raw__. Integer and double vectors are collectively known as
  __numeric__ vectors. 

1. __Lists__,  which are sometimes called recursive vectors because lists can 
  contain other lists. 

Differences between atomic vectors and lists: 

* Main: atomic vectors are __homogeneous__, while lists can be __heterogeneous__. 
* `NULL`

```{r}
c(NULL)
c(10, NULL, 23)

list(10, NULL, 23)

length(NULL)
length(c(NULL))
length(c(10, NULL, 23))
length(list(10, NULL, 23))
```

Every vector has two key properties: 

1.  Its __type__, which you can determine with `typeof()`.

```{r}
letters
typeof(letters)
typeof(1:10)
str(1:10)
```

2. Its __length__, which you can determine with `length()`.

```{r}
x <- list("a", "b", 1:10)
length(x)
```

Additional metadata: attributes. Gives __augmented vectors__ which build on additional behaviour. There are four important types of augmented vector:

* Factors are built on top of integer vectors.
* Dates and date-times are built on top of numeric vectors.
* Data frames and tibbles are built on top of lists.

This chapter will introduce you to these important vectors from simplest to most complicated.

## Important types of atomic vector

Focus on: logical, integer, double, and character.
Rarely used in data analysis: raw and complex.

### Logical

Values: `FALSE`, `TRUE`, and `NA`.

```{r}
1:10 %% 3 == 0
str(1:10 %% 3 == 0)

c(TRUE, TRUE, FALSE, NA)
```

### Numeric

Integer and double vectors are collectively: numeric vectors. 
In R, numbers are doubles by default. To make an integer, place an `L` after the number:

```{r}
typeof(1)
typeof(1L)
1.5L

typeof(1)
typeof(1:10)
typeof(c(1))

# Fun
1:1
typeof(1:1)
typeof(seq(1, 1))
seq(1, 1)
```

Distinction usually not important, but:

1. Doubles are approximations. 

```{r}
x <- sqrt(2)^2
x
x - 2
2 == sqrt(2)^2
identical(2, sqrt(2)^2)
all.equal(2, sqrt(2)^2)
dplyr::near(2, sqrt(2)^2)
dplyr::near
all.equal
lookup::lookup('all.equal')
lookup::lookup('all.equal.numeric')
```

2. Integers have one special value: `NA`. Doubles have four: `NA`, `NaN`, `Inf` and `-Inf`.

NA: Not available / missing (watch out dealing with abbreviations: North America, Namibia, Sodium, ...)
NaN: Not a number / not defined

All three special values `NaN`, `Inf` and `-Inf` can arise during division:

```{r}
c(-1, 0, 1) / 0
```

Use `is.finite()`, `is.infinite()`, and `is.nan()`:
    
    |                  |  0  | Inf | NA  | NaN |
    |------------------|-----|-----|-----|-----|
    | `is.finite()`    |  x  |     |     |     |
    | `is.infinite()`  |     |  x  |     |     |
    | `is.na()`        |     |     |  x  |  x  |
    | `is.nan()`       |     |     |     |  x  |

### Character


Character vectors: complex as each element of a character vector is a string, and a string can contain an arbitrary amount of data. 

stringr

R uses a global string pool (hash table): Each unique string is only stored in memory once, and every use of the string points to that representation:

```{r}
x <- "This is a reasonably long string."
pryr::object_size(x)

y <- rep(x, 1000)

1000*pryr::object_size(x)
pryr::object_size(y)
```

`y` doesn't take up 1,000x as much memory as `x`, because each element of `y` is just a pointer to that same string. A pointer is 8 bytes, so 1000 pointers to a 136 B string is 8 * 1000 + 136 = 8.13 kB.

### Missing values

Note that each type of atomic vector has its own missing value:

```{r}
NA            # logical
str(NA)
NA_integer_   # integer
NA_real_      # double
NA_character_ # character
```

Normally does not matter. But: type stability (__purrr__)

## Using atomic vectors

1.  How to convert from one type to another, and when that happens
    automatically.
2.  How to tell if an object is a specific type of vector.
3.  What happens when you work with vectors of different lengths.
4.  How to name the elements of a vector.
5.  How to pull out elements of interest.

### Coercion

There are two ways to convert, or coerce, one type of vector to another:

1.  Explicit coercion: `as.logical()`, `as.integer()`, `as.double()`, or `as.character()`. 
2.  Implicit coercion: when vector used in a specific context
    
Implicit coercion:

```{r}
x <- sample(20, 100, replace = TRUE)
typeof(x)
y <- x > 10
typeof(y)
sum(y)  # how many are greater than 10?
mean(y) # what proportion are greater than 10?
```

Implicit coercion in opposite direction (from integer to logical):

```{r, eval = FALSE}
if (length(x)) {
  # do something
}
```

Do not do this! Difficult to read. Instead be explicit: `length(x) > 0`.

What happens when creating a vector containing multiple types with `c()`: the most complex type always wins.

```{r}
typeof(c(TRUE, 1L))
typeof(c(1L, 1.5))
typeof(c(1.5, "a"))
c(1.5, "a")
```

...and it does so silently!

Multiple types in the same vector: use a list.

### Test functions

* `typeof()`: character.
* Want `TRUE` or `FALSE`. 
* Base R provides many functions like `is.vector()` and `is.atomic()`, but they often returns surprising results.
* Use `is_*` functions from __purrr__ summarised:

|                  | lgl | int | dbl | chr | list |
|------------------|-----|-----|-----|-----|------|
| `is_logical()`   |  x  |     |     |     |      |
| `is_integer()`   |     |  x  |     |     |      |
| `is_double()`    |     |     |  x  |     |      |
| `is_numeric()`   |     |  x  |  x  |     |      |
| `is_character()` |     |     |     |  x  |      |
| `is_atomic()`    |  x  |  x  |  x  |  x  |      |
| `is_list()`      |     |     |     |     |  x   |
| `is_vector()`    |  x  |  x  |  x  |  x  |  x   |

Each predicate also comes with a "scalar" version, like `is_scalar_atomic()`, which checks that the length is 1. This is useful, for example, if you want to check that an argument to your function is a single logical value.

### Scalars and recycling rules

* Implicitly coercing type
* Implicitly coerce the length of vectors; called vector __recycling__

```{r}
sample(10) + 100
runif(10) > 0.5

1:10 + 1:2
1:10 + 1:3
```

Vector recycling: can be used to create very succinct, clever code, it can also silently conceal problems.

Vectorised functions in __tidyverse__ will throw errors when you recycle anything other than a scalar. Use explicit recycling with `rep()`:

```{r, error = TRUE}
tibble(x = 1:4, y = 1:2)
tibble(x = 1:4, y = rep(1:2, 2))
tibble(x = 1:4, y = rep(1:2, each = 2))
```

### Naming vectors

```{r}
v = c(x = 1, y = 2, z = 4)
v
str(v)
```

Or after the fact with `purrr::set_names()`:

```{r}
set_names(1:3, c("a", "b", "c"))
```


### Subsetting

* `dplyr::filter()`: filter rows in a tibble
* `filter()` only works with tibble, so we'll need new tool for vectors: `[`. 
* `[` is the subsetting function, and is called like `x[a]`. 

Four types of things that you can subset a vector with:

1.  A numeric vector containing only integers. The integers must either be all 
    positive, all negative, or zero.
    
    Subsetting with positive integers keeps the elements at those positions:
    
    ```{r}
    x <- c("one", "two", "three", "four", "five")
    x[c(3, 2, 5)]
    ```
    
    By repeating a position, you can actually make a longer output than 
    input:
    
    ```{r}
    x[c(1, 1, 5, 5, 5, 2)]
    ```
    
    Negative values drop the elements at the specified positions:
    
    ```{r}
    x[c(-1, -3, -5)]
    -c(1, 3, 5)
    x[-c(1, 3, 5)]
    ```
    
    It's an error to mix positive and negative values:
    
    ```{r, error = TRUE}
    x[c(1, -1)]
    ```

    The error message mentions subsetting with zero, which returns no values:
    
    ```{r}
    x[0]
    ```
    
    This is not useful very often, but it can be helpful if you want to create 
    unusual data structures to test your functions with.
  
2.  Subsetting with a logical vector keeps all values corresponding to a
    `TRUE` value. This is most often useful in conjunction with the 
    comparison functions.
    
    ```{r}
    x <- c(10, 3, NA, 5, 8, 1, NA)
    
    # All non-missing values of x
    x[!is.na(x)]
    
    # All even (or missing!) values of x
    x[x %% 2 == 0]
    ```

3.  If you have a named vector, you can subset it with a character vector:
    
    ```{r}
    x <- c(abc = 1, def = 2, xyz = 5)
    x[c("xyz", "def")]
    ```
    
    Like with positive integers, you can also use a character vector to 
    duplicate individual entries.

4.  The simplest type of subsetting is nothing, `x[]`.
    Not really useful for subsetting vectors.
    Subsetting matrices (or arrays), e.g if `x` is 2d, `x[1, ]` selects the first row and 
    all the columns, and `x[, -1]` selects all rows and all columns except
    the first. Note `x[1, ]` vs `x[1, , drop = FALSE]`
    
```{r}
(1:10)[]

x = matrix(1:4, 2, 2)
x
x[1, ]
x[1, , drop = FALSE]
dim(x)
dim(x[1, ])
dim(x[1, , drop = FALSE])
```


Variation of `[`: `[[`:

* `[[` only ever extracts a single element, and always drops names.
* Whenever you want to make it clear that you're extracting a single item, as in a for loop. 
* The distinction between `[` and `[[` is most important for lists, as we'll see shortly.

## Recursive vectors (lists) {#lists}

Lists are a step up in complexity from atomic vectors, because lists can contain other lists. This makes them suitable for representing hierarchical or tree-like structures. You create a list with `list()`:

```{r}
x <- list(1, 2, 3)
x
```

A very useful tool for working with lists is `str()` because it focusses on the **str**ucture, not the contents.

```{r}
str(x)

x_named <- list(a = 1, b = 2, c = 3)
str(x_named)
```

Unlike atomic vectors, `list()` can contain a mix of objects:

```{r}
y <- list("a", 1L, 1.5, TRUE)
str(y)
```

Lists can even contain other lists!

```{r}
z <- list(list(1, 2), list(3, 4))
str(z)
str(z, 1)
```

### Visualising lists

To explain more complicated list manipulation functions, it's helpful to have a visual representation of lists. For example, take these three lists:

```{r}
x1 <- list(c(1, 2), c(3, 4))
x2 <- list(list(1, 2), list(3, 4))
x3 <- list(1, list(2, list(3)))

str(x1)
str(x2)
str(x3)

str(x3, 1)
```

### Subsetting

There are three ways to subset a list, which I'll illustrate with a list named `a`:

```{r}
a <- list(a = 1:3, b = "a string", c = pi, d = list(-1, -5))
```

*   `[` extracts a sub-list. The result will always be a list.

    ```{r}
    str(a[1:2])
    str(a[4])
    ```
    
    Like with vectors, you can subset with a logical, integer, or character
    vector.
    
*   `[[` extracts a single component from a list. It removes a level of 
    hierarchy from the list.

    ```{r}
    a
    str(a)
    str(a[[1]])
    str(a[1])
    str(a[[4]])
    ```

*   `$` is a shorthand for extracting named elements of a list. It works
    similarly to `[[` except that you don't need to use quotes.
    
    ```{r}
    a$a
    a[["a"]]
    ```

The distinction between `[` and `[[` is really important for lists, because `[[` drills down into the list while `[` returns a new, smaller list.

## Attributes

Any vector can contain arbitrary additional metadata through its __attributes__. You can think of attributes as named list of vectors that can be attached to any object. 
You can get and set individual attribute values with `attr()` or see them all at once with `attributes()`.

```{r}
x <- 1:10
attr(x, "greeting")
attr(x, "greeting") <- "Hi!"
x
attr(x, "farewell") <- "Bye!"
attributes(x)
x
```

There are three very important attributes that are used to implement fundamental parts of R:

1. __Names__ are used to name the elements of a vector.
1. __Dimensions__ (dims, for short) make a vector behave like a matrix or array.
1. __Class__ is used to implement the S3 object oriented system.

* Names above
* Dimensions (`matrix` / `array`)
* Class: controls how __generic functions__ work

Generic functions are key to object oriented programming in R-
Dynamic dispatch.

Details of object oriented programming in _Advanced R_ at <http://adv-r.had.co.nz/OO-essentials.html#s3>.

Here's what a typical generic function looks like:

```{r}
as.Date
methods("as.Date")
```

For example, if `x` is a character vector, `as.Date()` will call `as.Date.character()`; if it's a factor, it'll call `as.Date.factor()`.

You can see the specific implementation of a method with `getS3method()`:

```{r}
getS3method("as.Date", "default")
getS3method("as.Date", "numeric")
```

The most important S3 generic is `print()`: it controls how the object is printed when you type its name at the console.

```{r}
methods("print")
```

Other important generics are the subsetting functions `[`, `[[`, and `$`. 

## Augmented vectors

Vectors with additional __attributes__, including class:

* Factors
* Dates 
* Date-times
* Tibbles

These are described below.

### Factors

Represent categorical data; built on top of integers, and have a levels attribute:

```{r}
x <- factor(c("ab", "cd", "ab"), levels = c("ab", "cd", "ef"))
typeof(x)
attributes(x)
```

### Dates and date-times

Dates in R are numeric vectors that represent the number of days since 1 January 1970.

```{r}
x <- as.Date("1971-01-01")
unclass(x)

typeof(x)
attributes(x)
```

Date-times are numeric vectors with class `POSIXct` that represent the number of seconds since 1 January 1970. (In case you were wondering, "POSIXct" stands for "Portable Operating System Interface", calendar time.)

```{r}
x <- lubridate::ymd_hm("1970-01-01 01:00")
unclass(x)

typeof(x)
attributes(x)
```

The `tzone` attribute is optional. It controls how the time is printed, not what absolute time it refers to.

```{r}
attr(x, "tzone") <- "US/Pacific"
x

attr(x, "tzone") <- "US/Eastern"
x
```

### Tibbles

Tibbles are augmented lists: they have class "tbl_df" + "tbl" + "data.frame", and `names` (column) and `row.names` attributes:

```{r}
tb <- tibble::tibble(x = 1:5, y = 5:1)
typeof(tb)
attributes(tb)
```

The difference between a tibble and a list is that all the elements of a data frame must be vectors with the same length. All functions that work with tibbles enforce this constraint.

Traditional data.frames have a very similar structure:

```{r}
df <- data.frame(x = 1:5, y = 5:1)
typeof(df)
attributes(df)
```

The main difference is the class. The class of tibble includes "data.frame" which means tibbles inherit the regular data frame behaviour by default.

