---
output: html_document
---

# Functions

* Automate common tasks: more powerful and general than copy-and-pasting (easier to debug, improve, ...).
* Three big advantages over using copy-and-paste:

1. Name: makes your code easier to understand.
2. Only need to update code in one place.
3. Risk of incidental mistakes when copy'n'paste.

* Difficult
* Rarely/never only one true implementation.
* Here, some tips; not exhaustive guide.

## When should you write a function?

* Always
* But maybe not to begin with

R4DS:

* "You should consider writing a function whenever you've copied and pasted a block of code more than twice (i.e. you now have three copies of the same code)."
* But also if the code is just being run once:
    + Put functionality in one file `include-functions.R` and in `analysis.R` simply call the function.
    
For example, take a look at this code. What does it do?

```{r}
df <- tibble::tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)

df$a <- (df$a - min(df$a, na.rm = TRUE)) / 
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$b <- (df$b - min(df$b, na.rm = TRUE)) / 
  (max(df$b, na.rm = TRUE) - min(df$a, na.rm = TRUE))
df$c <- (df$c - min(df$c, na.rm = TRUE)) / 
  (max(df$c, na.rm = TRUE) - min(df$c, na.rm = TRUE))
df$d <- (df$d - min(df$d, na.rm = TRUE)) / 
  (max(df$d, na.rm = TRUE) - min(df$d, na.rm = TRUE))
```












* Rescales each column to have a range from 0 to 1.
* Did you spot the mistake? (In code for `df$b`: `a` is still present.)

Convert into function. Analyse the code:

* How many inputs does it have?

```{r, eval = FALSE}
(df$a - min(df$a, na.rm = TRUE)) /
  (max(df$a, na.rm = TRUE) - min(df$a, na.rm = TRUE))
```

Answer: One input: `df$a`. (But a vector.)

Rewrite the code using temporary variables with general names; let's call the input `x`:

```{r}
x <- df$a
(x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))

rescale01_ver0 = function(x) {
  (x - min(x, na.rm = TRUE)) / (max(x, na.rm = TRUE) - min(x, na.rm = TRUE))
}
rescale01_ver0(df$a)
rescale01_ver0(df$b)
rescale01_ver0(df$c)
rescale01_ver0(df$d)
```

Optimization:

* Duplication: Computing `min`/`max` three times, two times sufficient:

```{r}
rng <- range(x, na.rm = TRUE)
(x - rng[1]) / (rng[2] - rng[1])
```

* Pulling out intermediate calculations into named variables is a good practice! More clear what the code is doing.

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(c(0, 5, 10))

rescale01_ver0(c(0, 5, 10))

identical(rescale01_ver0(c(0, 5, 10)), rescale01(c(0, 5, 10))) # exactly equal
all.equal(rescale01_ver0(c(0, 5, 10)), rescale01(c(0, 5, 10))) # numerically equal

identical(1e-20, 2e-20)
all.equal(1e-20, 2e-20)

all.equal(1e-6, 2e-6)
all.equal(1e-6, 2e-6, tolerance = 1e-2)
```



Three key steps to creating a new function:

1. Pick a __name__ for the function. (E.g. `rescale01`.)
2. Identify the inputs, or __arguments__. The API. (E.g. `function(x)`, `function(x, y, z)`, ...)
3. Code you in __body__ of the function. (A `{` block that immediately follows `function(...)`.)


```{r}
test_func = function(x, y) {
  200*x + y
}

test_func(10, 2)
test_func(x = 10, y = 2)
test_func(y = 2, x = 10)
```

Fourth:

* Testing (many possibilities)
    + Version 1: Maybe long and inefficient, but answer is verified.
    + Version 2: Unit test (__testthat__ package) / static analysis (__lintr__) / regression test (API, the whole package)
        - <http://r-pkgs.had.co.nz/tests.html>

At this point it's a good idea to check your function with a few different inputs:

```{r}
rescale01(c(-10, 0, 10))
rescale01(c(1, 2, 3, NA, 5))
```

We can simplify the original example now that we have a function:

```{r}
df$a <- rescale01(df$a)
df$b <- rescale01(df$b)
df$c <- rescale01(df$c)
df$d <- rescale01(df$d)
```

* Compared to the original: easier to understand, no copy-and-paste errors.
* (Some duplication: doing the same thing to multiple columns.)
* Advantage: if requirements change, change in one place. E.g. variables include infinite values, and `rescale01()` fails:

```{r}
x <- c(1:10, Inf)
rescale01(x)
```

Because we've extracted the code into a function, we only need to make the fix in one place:

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE, finite = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}
rescale01(x)
```

* "Do not repeat yourself" (or DRY) principle: The more repetition you have in your code, the more places you need to remember to update when things change (and they always do!), and the more likely you are to create bugs over time.



### Practice

R4DS Exercises p. 273 (online 19.2.1 Practice): exercise 6:

* What do the following functions do? Why are they useful even though they are so short?

```{r}
is_directory <- function(x) file.info(x)$isdir
is_readable <- function(x) file.access(x, 4) == 0
```


## Functions are for humans and computers

* Functions are not just for the computer, but are also for humans.
* R doesn't care what your function is called, or what comments it contains, but these are important for human readers.

Advices for helping humans read the code:

* Name of a function is important: Short, but clearly evoke what the function does. Hard! Better to be clear than short: IDE such as RStudio has autocompletion.
* Function names should be verbs, and arguments should be nouns. Exceptions: 
    + nouns are ok if the function computes a very well known noun (i.e. `mean()` is better than `compute_mean()`)
    + accessing some property of an object (i.e. `coef()` is better than `get_coefficients()`)
    + A good sign that a noun might be a better choice is if you're using a very broad verb like "get", "compute", "calculate", or "determine". Use your best judgement and don't be afraid to rename a function if you figure out a better name later. !!!!


```{r, eval = FALSE}
# Too short
f()

# Not a verb, or descriptive
my_awesome_function()

# Long, but clear
impute_missing()
collapse_years()
```

* Multiple words: 
    + "snake\_case": each lowercase word is separated by an underscore.
    + "camelCase" is a popular alternative.

Choice is a matter of personal opinion. R4DS: "Important thing is to be consistent: pick one or the other and stick with it." (Until you change your mind.)

Old projects tends to change over time: R itself is not very consistent. Difficult.

```{r, eval = FALSE}
# Never do this!
col_mins <- function(x, y) {}
rowMaxes <- function(y, x) {}
```

Family of functions that do similar things: consistent names and arguments.
Use a common prefix to indicate that they are connected. Easier for autocomplete.

```{r, eval = FALSE}
# Good
input_select()
input_checkbox()
input_text()

# Not so good
select_input()
checkbox_input()
text_input()
```

An example: `stringr` package prefix `str_` helps autocomplete.

But really a matter of personal taste -- and writing vs reading.

Avoid overriding names! Impossible to do in general, but avoiding the most common names from base R:

```{r, eval = FALSE}
# Don't do this!
T <- FALSE
c <- 10
mean <- function(x) sum(x)
```

## Comments

```{r, eval = FALSE}
# Load data --------------------------------------

# Plot data --------------------------------------
```

## Conditional execution

An `if` statement allows you to conditionally execute code. It looks like this:

```{r, eval = FALSE}
if (condition) {
  # code executed when condition is TRUE
} else {
  # code executed when condition is FALSE
}
```

### Conditions

`condition` must evaluate to either `TRUE` or `FALSE`.

If it's a vector, you'll get a warning message; if it's an `NA`, you'll get an error. Watch out for these messages in your own code:

```{r, error = TRUE}
if (c(TRUE, FALSE)) {}

if (NA) {}
```

Multiple logical expressions: `||` (or) and `&&` (and).

"Short-circuiting":

* `||`: sees the first `TRUE` it returns `TRUE` without computing anything else.
* `&&`: sees the first `FALSE` it returns `FALSE`. 

Vectorized:

* `|` 
* `&`
* `==`
* `any()` or `all()`

```{r}
c(TRUE, FALSE) | c(TRUE, TRUE)
c(FALSE, FALSE) || c(FALSE, TRUE) # Only first element checked!
c(FALSE, FALSE) == c(FALSE, TRUE)
```

### Multiple conditions

You can chain multiple if statements together:

```{r, eval = FALSE}
if (this) {
  # do that
} else if (that) {
  # do something else
} else {
  # 
}
```


## Check preconditions

It's good practice to check important preconditions, and throw an error (with `stop()`), if they are not true:

```{r}
wt_mean <- function(x, w) {
  if (length(x) != length(w)) {
    stop("`x` and `w` must be the same length")
  }
  sum(w * x) / sum(w)
}
```

* Tradeoff:

```{r}
wt_mean <- function(x, w, na.rm = FALSE) {
  if (!is.logical(na.rm)) {
    stop("`na.rm` must be logical")
  }
  if (length(na.rm) != 1) {
    stop("`na.rm` must be length 1")
  }
  if (length(x) != length(w)) {
    stop("`x` and `w` must be the same length", call. = FALSE)
  }
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}
```

```{r, error = TRUE}
wt_mean <- function(x, w, na.rm = FALSE) {
  stopifnot(is.logical(na.rm), length(na.rm) == 1)
  stopifnot(length(x) == length(w))
  
  if (na.rm) {
    miss <- is.na(x) | is.na(w)
    x <- x[!miss]
    w <- w[!miss]
  }
  sum(w * x) / sum(w)
}
wt_mean(1:6, 6:1, na.rm = "foo")
```

Note that when using `stopifnot()` you assert what should be true rather than checking for what might be wrong.

### Variadic arguments / dot-dot-dot (...)

Many functions in R take an arbitrary number of inputs:

```{r}
sum(1, 2, 3, 4, 5, 6, 7, 8, 9, 10)
stringr::str_c("a", "b", "c", "d", "e", "f")
```


Special -- variadic -- argument: `...` (pronounced dot-dot-dot). Captures any number of arguments that aren't otherwise matched. 

Send those `...` on to another function:

```{r}
commas <- function(...) stringr::str_c(..., collapse = ", ")
commas(letters[1:10])

rule <- function(..., pad = "-") {
  title <- paste0(...)
  width <- getOption("width") - nchar(title) - 5
  cat(title, " ", stringr::str_dup(pad, width), "\n", sep = "")
}
rule("Important output")
```

Price: any misspelled arguments will not raise an error. This makes it easy for typos to go unnoticed:

```{r}
x <- c(1, 2, NA)
sum(x, na.mr = TRUE)
```

If you just want to capture the values of the `...`, use `list(...)`.

```{r}
capt = function(...) {
  x = list(...)
  print(x)
  return(invisible(x))
}

x = capt(1, 2, 3)
str(x)
```


### Lazy evaluation


Arguments in R are lazily evaluated: they're not computed until they're needed. That means if they're never used, they're never called. This is an important property of R as a programming language, but is generally not important when you're writing your own functions for data analysis. 

```{r}
f <- function(x) {
  10
}
f(stop("This is an error!"))
```

```{r, error=TRUE}
f <- function(x) {
  force(x)
  10
}
f(stop("This is an error!"))
```


You can read more about lazy evaluation at <http://adv-r.had.co.nz/Functions.html#lazy-evaluation>.


## Return values

Figuring out what your function should return is usually straightforward: it's why you created the function in the first place! There are two things you should consider when returning a value: 

1. Does returning early make your function easier to read? 
2. Can you make your function pipeable?

### Explicit return statements

The value returned by the function is usually the last statement it evaluates, but you can choose to return early by using `return()`. I think it's best to save the use of `return()` to signal that you can return early with a simpler solution. A common reason to do this is because the inputs are empty:

```{r}
complicated_function <- function(x, y, z) {
  if (length(x) == 0 || length(y) == 0) {
    return(0)
  }
    
  # Complicated code here
}

```

Another reason is because you have a `if` statement with one complex block and one simple block. For example, you might write an if statement like this:

```{r, eval = FALSE}
f <- function() {
  if (x) {
    # Do 
    # something
    # that
    # takes
    # many
    # lines
    # to
    # express
  } else {
    # return something short
  }
}
```

But if the first block is very long, by the time you get to the `else`, you've forgotten the `condition`. One way to rewrite it is to use an early return for the simple case:

```{r, eval = FALSE}
f <- function() {
  if (!x) {
    return(something_short)
  }

  # Do 
  # something
  # that
  # takes
  # many
  # lines
  # to
  # express
}
```

This tends to make the code easier to understand, because you don't need quite so much context to understand it.


### Type stability

```{r}
f <- function(x) {
  if (!x) {
    return(0)
  }
  
  return("foo")
}
```


### Writing pipeable functions

* Think about the return value.
* For highly flexible pipeline: input type = output type (__dplyr__: data frame). 



## Environment

The last component of a function is its environment. Example:

```{r}
f <- function(x) {
  x + y
} 
```

In many programming languages, this would be an error, because `y` is not defined inside the function. 

R uses __lexical scoping__ (__static scope__) to find the value associated with a name. Since `y` is not defined inside the function, R will look in the __environment__ where the function was defined:

```{r}
y <- 100
f(10)

y <- 1000
f(10)
```

Advantage: allows R to be very consistent. Every name is looked up using the same set of rules.

Recipe for bugs? Yes, if you really want:

```{r}
`+` <- function(x, y) {
  if (runif(1) < 0.1) {
    sum(x, y)
  } else {
    sum(x, y) * 1.1
  }
}
table(replicate(1000, 1 + 2))

`+` <- function(x, y) {
  #x + y + runif(1)
  sum(x, y, runif(1))
}
1+2
1+2
1+2

rm(`+`)
```

## Side effects

```{r}
f = function(d) {
  plot(d)
  return(invisible(d))
  #return(invisble(NULL))
  #return(invisble(TRUE))
}
r = f(1:3)
```



## References

More in R4DS and in [_Advanced R_](http://adv-r.had.co.nz).

