---
title: "Iterations (Chapter 17 - former part)"
author: "Torben"
date: "December 15, 2017"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, comment = NA)
```

```{r}
library(tidyverse)
theme_set(theme_bw())
```

# Simple iterations

## For loops

```{r}
df <- tibble(
  a = rnorm(10),
  b = rnorm(10),
  c = rnorm(10),
  d = rnorm(10)
)
```

Where we wants to call the `summary` function on each column.

```{r}
summary(df$a)
summary(df$b)
summary(df$c)
summary(df$d)
```

However, there is (several!) a more efficient way

```{r}
output <- vector(length = ncol(df), mode = "list")
for (i in seq_along(df)) {
  output[[i]] <- summary(df[[i]])
}
```

And the `output` holds the results:
```{r, echo=FALSE}
output
```

For these simple `for` loops, R is running though an object of indicies, `X`, by the construction `for(i in X)`.
The use of `seq_along` ensures that border cases of `X` being `NULL` (or empty) we obtain `integer(0)`, 
whereas the simpler `1:length(X)` would fail in such an instance.
```{r}
y <- vector("double", 0)
seq_along(y)
1:length(y)
```

### Exercises

1, 2, 4

## More on for loops

We can also modify the elements in a `data.frame` or `tibble`
```{r}
center_std <- function(x) (x-mean(x))/sd(x)

x <- 1:10
mean(x) 
sd(x)
mean(center_std(x))
sd(center_std(x))
```

For the columns in the `df` we can center and standardise by 
```{r}
df_ <- df
for (i in seq_along(df_)) {
  df_[[i]] <- center_std(df_[[i]])
}
df_
```

We may also loop over names
```{r}
df_n <- df
for (i in names(df_n)) {
  df_n[[i]] <- center_std(df_n[[i]])
}
df_n
```

### The flexibility of lists

The book's example on using lists for temporal storage of results is relevant enough to go through.
```{r}
means <- c(0, 1, 2)                          ## 3 different mean values

output <- double()                           ## The output container
for (i in seq_along(means)) {                ## loop over the 'means' vector
  n <- sample(100, 1)                        ## randomly 'sample' 1 number between 1 and 100
  output <- c(output, rnorm(n, means[[i]]))  ## concatenate the result to the output
}
str(output)                                  ## class of output
```

Here we dynamically allocates memory for `output` - this is expensive and should be avoided.

Using lists, we let R handle this allocation *per* constructed list element (this is cheaper)
```{r}
out <- vector("list", length(means))
for (i in seq_along(means)) {
  n <- sample(100, 1)
  out[[i]] <- rnorm(n, means[[i]])
}
str(out)
str(unlist(out))
```

### Exericses

1, 2, 3, (4)

## Functionals

In order for us to compute the `mean` of each column in a `data.frame` we can use the `for` loop constructions above. However, it is much better to write a function that takes care of all the bookkeeping once.

```{r}
col_mean <- function(df, ...) { ## The ... allows us to pass additional arguments to inner functions
  output <- rep.int(NA, ncol(df))
  for (i in seq_along(df)) output[i] <- mean(df[[i]], ...) ## eg. trim or na.rm
  output
}

col_mean(df)
df$b[4] <- NA
col_mean(df, na.rm = TRUE)
```

We can however, make us of an even more flexible contruction, where the `mean` can be replaced by other functions.

```{r}
col_fun <- function(df, fun, ...) {
  out <- replicate(ncol(df), NULL) 
  for (i in seq_along(df)) out[[i]] <- fun(df[[i]], ...)
  out
}

col_fun(df, summary)
col_fun(df, median, na.rm = TRUE)
```

### Exercise
1, 2