---
output: html_document
---

# Many models

## Introduction

1. Many simple models: better understand complex datasets.

2. List-columns (complex columns) to store arbitrary data structures in a data frame.
   E.g. a linear model.
   
3. __broom__: tidying model information

Data: life expectancy around the world

### Prerequisites

```{r setup, message = FALSE}
library(modelr)
library(tidyverse)
```

## gapminder


* BBC: <https://www.youtube.com/watch?v=jbkSRLYSojo>
* The Best Stats You've Ever Seen | Hans Rosling | TED Talks: <https://www.youtube.com/watch?v=y1kg5k6_fcA>

Hans Rosling (1948-2017), a Swedish doctor and statistician-

gapminder data: progression of countries over time:

```{r}
library(gapminder)
gapminder
```

"How does life expectancy (`lifeExp`) change over time (`year`) for each country (`country`)?".

```{r}
gapminder %>% 
  ggplot(aes(year, lifeExp, group = country)) +
    geom_line(alpha = 1/3)
```

Difficult to see what's going on. Small dataset.

General: life expectancy steadily improving. Exceptions.

Strong signal (overall linear growth).

```{r, out.width = "33%", fig.asp = 1, fig.width = 3, fig.align='default'}
nz <- filter(gapminder, country == "New Zealand")
nz %>% 
  ggplot(aes(year, lifeExp)) + 
  geom_line() + 
  ggtitle("Full data = ")

nz_mod <- lm(lifeExp ~ year, data = nz)
nz %>% 
  add_predictions(nz_mod) %>%
  ggplot(aes(year, pred)) + 
  geom_line() + 
  #geom_line(aes(year, lifeExp), data = nz, linetype = "dashed") + 
  ggtitle("Linear trend + ")

nz %>% 
  add_residuals(nz_mod) %>% 
  ggplot(aes(year, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line() + 
  ggtitle("Residuals")
```

How can we easily fit that model to every country?

### Nested data

Copy and pasting that code multiple times?

Make function + purrr. 

Not for each variable but a subset of rows: __nested data frame__.

```{r}
by_country <- gapminder %>% 
  group_by(country, continent) %>% 
  nest()

by_country
```

Grouping on both `continent` and `country`: Given `country`, `continent` is fixed
Easy way to carry an extra variable along for the ride.

`data`: is a list of data frames (or tibbles, to be precise).
`str()`: very long output
Difficult to look at. (Work in progress.)

```{r}
by_country$data[[1]]
by_country$data[1:2]
```

Grouped data frame vs nested data frame: 

  * Grouped: each row is an observation
  * Nested: each row is a group / meta-observation

### List-columns

```{r}
country_model <- function(df) {
  lm(lifeExp ~ year, data = df)
}
```

Apply it to every data frame: List -> use `purrr::map()` to apply `country_model` to each element:

```{r}
models <- map(by_country$data, country_model)
```

`models`: Current a free-floating object. 

```{r}
by_country <- by_country %>% 
  mutate(model = map(data, country_model))
by_country
```

All the related objects are stored together!!

```{r}
by_country %>% 
  filter(continent == "Europe")
by_country %>% 
  arrange(continent, country)
```

Filtering, arranging etc. works and data is consistent!

### Unnesting

Previously we computed the residuals of a single model with a single dataset. Now we have 142 data frames and 142 models. To compute the residuals, we need to call `add_residuals()` with each model-data pair:

```{r}
by_country <- by_country %>% 
  mutate(
    resids = map2(data, model, add_residuals)
  )
by_country
```

# `map`:  `map(.x, .f, ...) ~ list(.f(.x[[1]]), .f(.x[[2]]), ...)`
# `map2`: `map(.x, .y, .f, ...) ~ list(.f(.x[[1]], .y[[1]]), .f(.x[[2]], .y[[2]]), ...)`

Plotting? Turn list of data frames back into a regular data frame. 

`nest()`: regular data frame -> nested data frame
`unnest()`: nested data frame -> regular data frame

```{r}
resids <- unnest(by_country, resids) # keeps all std columns, unnest only `resids`
resids
```

Note: each regular column is repeated one for each row in the nested column.

Regular data frame -> plot the residuals:

```{r}
resids %>% 
  ggplot(aes(year, resid)) +
    geom_line(aes(group = country), alpha = 1/3) + 
    geom_smooth(se = FALSE)

if (FALSE) {
  # WRONG!
  resids %>% 
  ggplot(aes(year, resid, group = country)) +
    geom_line(alpha = 1/3) + 
    geom_smooth(se = FALSE)
  # 
  # # loess
  # resids %>% 
  # ggplot(aes(year, resid)) +
  #   geom_line(aes(group = country), alpha = 1/3) + 
  #   geom_smooth(method = "loess", se = FALSE)
  # 
  # resids %>% 
  # ggplot(aes(year, resid)) +
  #   geom_line(aes(group = country), alpha = 1/3) + 
  #   geom_smooth(method = "gam", se = FALSE)
  # 
  # resids %>% 
  # ggplot(aes(year, resid)) +
  #   geom_line(aes(group = country), alpha = 1/3) + 
  #   geom_smooth(aes(color = "LOESS"), method = "loess", se = FALSE) + 
  #   geom_smooth(aes(color = "GAM"), method = "gam", se = FALSE)
}
```

Facetting by continent is particularly revealing:
 
```{r}
resids %>% 
  ggplot(aes(year, resid, group = country)) +
    geom_line(alpha = 1/3) + 
    facet_wrap(~ continent)

resids %>% 
  ggplot(aes(year, resid)) +
    geom_line(aes(group = country), alpha = 1/3) + 
    geom_smooth(se = FALSE) + # !
    facet_wrap(~ continent)
```

Missed small patterns. 

Africa: some very large residuals.

### Model quality

Residuals vs general measurements of model quality.

__broom__ package: a general set of functions to turn (information about) models into tidy data. 

`broom::glance()`:

```{r}
broom::glance(nz_mod)
```

We can use `mutate()` and `unnest()` to create a data frame with a row for each country:

```{r}
by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance)
```

Still includes all list columns. 

Default behaviour of `unnest()`. To suppress these columns we use `.drop = TRUE`:

```{r}
glance <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance, .drop = TRUE)
glance
```

(Not all variables are printed: there's a lot of useful stuff there.)

```{r}
glance %>% 
  arrange(r.squared)
```

Africa.

```{r}
glance %>% 
  ggplot(aes(continent, r.squared)) + 
    geom_jitter(width = 0.5)

glance %>% 
  ggplot(aes(continent, r.squared, color = continent)) + 
    geom_jitter(width = 0.25, height = 0)
```

Countries with particularly bad $R^2$:

```{r}
bad_fit <- filter(glance, r.squared < 0.25)

gapminder %>% 
  semi_join(bad_fit, by = "country") %>% 
  ggplot(aes(year, lifeExp, colour = country)) +
    geom_line()
```

We see two main effects here: the tragedies of the HIV/AIDS epidemic and the Rwandan genocide.

### Exercises

R4DS Exercises p. 409 (online 25.2.5 Exercises).

## List-columns

```{r}
data.frame(x = list(1:3, 3:5))
```

You can prevent `data.frame()` from doing this with `I()`, but the result doesn't print particularly well:

```{r}
data.frame(
  x = I(list(1:3, 3:5)), 
  y = c("1, 2", "3, 4, 5")
)
```

Tibble alleviates this problem by being lazier (`tibble()` doesn't modify its inputs) and by providing a better print method:

```{r}
tibble(
  x = list(1:3, 3:5), 
  y = c("1, 2", "3, 4, 5")
)
```

It's even easier with `tribble()` as it can automatically work out that you need a list:

```{r}
tribble(
   ~x, ~y,
  1:3, "1, 2",
  3:5, "3, 4, 5"
)
```

Often as intermediate data structure. 


Generally there are three parts of an effective list-column pipeline:

1.  Create: 
    * `nest()`
    * `summarise()` + `list()`
    * `mutate()` + a map function, as above

2.  Transforming existing list columns with `map()`, `map2()` or `pmap()`. 
    
3.  `unnest()`

## Creating list-columns

Rarely create list-columns manually with `tibble()`.

General advice: make sure they're homogeneous: each element should contain the same type of thing.

### With nesting

`nest()`

1. Grouped data frame. Keeps grouping, rest into the list-column:

```{r}
gapminder

gapminder %>% 
  group_by(country, continent) %>% 
  nest()
```

2. Specifying which columns you want to nest:

```{r}
gapminder %>% 
  nest(year:gdpPercap)
```

### From vectorised functions

Some functions: atomic vector -> list

`stringr::str_split()`: character vector -> list of character vectors:

```{r}
df <- tribble(
  ~x1,
  "a,b,c", 
  "d,e,f,g"
) 

df %>% 
  mutate(x2 = stringr::str_split(x1, ","))
```

`unnest()` knows how to handle these lists of vectors:

```{r}
df %>% 
  mutate(x2 = stringr::str_split(x1, ",")) %>% 
  unnest()
```

(Exactly this patern: `tidyr:separate_rows()`.)

`map()`, `map2()`, `pmap()` from __purrr__: 

```{r}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = -1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)

sim

sim %>%
  mutate(sims = invoke_map(f, params, n = 10))
```


### From multivalued summaries

`summarise()`: must return a single value. `quantile()`!

```{r, error = TRUE}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = quantile(mpg))
```

Wrap in a list! 

```{r}
mtcars %>% 
  group_by(cyl) %>% 
  summarise(q = list(quantile(mpg)))
```

To make useful results with unnest, you'll also need to capture the probabilities:

```{r}
probs <- c(0.01, 0.25, 0.5, 0.75, 0.99)
mtcars %>% 
  group_by(cyl) %>% 
  summarise(p = list(probs), q = list(quantile(mpg, probs))) %>% 
  unnest()
```

### From a named list

```{r}
x <- list(
  a = 1:5,
  b = 3:4, 
  c = 5:6
) 
x

df <- enframe(x)
df
```


## Simplifying list-columns

Often: Simplify list-column to regular column:

1.  Single value: `mutate()` with `map_lgl()`, `map_int()`, `map_dbl()`, and `map_chr()` to create an atomic vector.
    
2.  Many values: `unnest()` to convert list-columns back  to regular columns, repeating the rows as many times as necessary.

### List to vector


```{r}
df <- tribble(
  ~x,
  letters[1:5],
  1:3,
  runif(5)
)

df
  
df %>% mutate(
  type = map_chr(x, typeof),
  length = map_int(x, length)
)
```


### Unnesting

`unnest()`:

```{r}
df = tibble(x = 1:2, y = list(1:4, 1))
df
df %>% unnest(y)
```

This means that you can't simultaneously unnest two columns that contain different number of elements:

```{r, error = TRUE}
# Ok, because y and z have the same number of elements in
# every row
df1 <- tribble(
  ~x, ~y,           ~z,
   1, c("a", "b"), 1:2,
   2, "c",           3
)
df1
df1 %>% unnest(y, z)
df1 %>% unnest()

# Doesn't work because y and z have different number of elements
df2 <- tribble(
  ~x, ~y,           ~z,
   1, "a",         1:2,  
   2, c("b", "c"),   3
)
df2
df2 %>% unnest(y, z)
```


## Making tidy data with broom

__broom__ package:

1.  `broom::glance(model)`: a row for each model with model summary
   
2.  `broom::tidy(model)`: a row for each coefficient in the model (estimate, variance, ...)
    
3.  `broom::augment(model, data)` a row for each row in `data` with extra values like residuals, and influence statistics.

