---
output: html_document
---

# Pipes

## Introduction

* Pipes are a powerful tool for clearly expressing a sequence of multiple operations.
* Now, more detail on how they work.

### Prerequisites

* `%>%` comes from the __magrittr__ package by Stefan Milton Bache.
* Packages in the __tidyverse__ load `%>%` automatically.
* Here, focus on piping:

```{r setup, message = FALSE}
library(magrittr)
```

Pipe: help write/read code.

```{r}
sum(sqrt(1:10))
1:10 %>% sqrt() %>% sum()

dplyr::mutate(diamonds, price_per_carat = price / carat)

diamonds %>% 
  dplyr::mutate(price_per_carat = price / carat)
```

* Worry: creates many copies of your data and takes up a lot of memory. NO!
* Proactively worrying about memory is not a useful way to spend your time: worry about it when it becomes a problem (i.e. you run out of memory), not before.
* R isn't stupid, and it will share columns across data frames, where possible.

```{r}
diamonds <- ggplot2::diamonds
diamonds2 <- diamonds %>% 
  dplyr::mutate(price_per_carat = price / carat)

diamonds %>% 
  dplyr::filter(carat >= 2) %>% 
  dplyr::mutate(price_per_carat = price / carat)

pryr::object_size(diamonds)
pryr::object_size(diamonds2)
pryr::object_size(diamonds, diamonds2)
```

`pryr::object_size()` gives the memory occupied by all of its arguments. The results seem counter-intuitive at first:

* `diamonds` takes up 3.46 MB,
* `diamonds2` takes up 3.89 MB,
* `diamonds` and `diamonds2` together take up 3.89 MB!

`diamonds2` has 10 columns in common with `diamonds`: no need to duplicate all that data
* Only copied if modified

```{r}
diamonds$carat[1] <- NA
pryr::object_size(diamonds)
pryr::object_size(diamonds2)
pryr::object_size(diamonds, diamonds2)
```

(Note that we use `pryr::object_size()` here, not the built-in `object.size()`. 
`object.size()` only takes a single object so it can't compute how data is shared across multiple objects.)

The pipe works by performing a "lexical transformation": behind the scenes, __magrittr__ reassembles the code in the pipe to a form that works by overwriting an intermediate object. When you run a pipe like the one above, magrittr does something like this:

```{r, eval = FALSE}
my_pipe <- function(.) {
  dplyr::mutate(., price_per_carat = price / carat)
}
my_pipe(diamonds)
```

This means that the pipe won't work for two classes of functions:

1. Functions that use the current environment. For example, `assign()`
   will create a new variable with the given name in the current environment:
     
```{r}
assign("x", 10)
x

# assign("x", 100) !==
"x" %>% assign(100)
x

my_pipe <- function(., n) {
  assign(., n)
  x
}
my_pipe("x", 100)
x
```

If you do want to use assign with the pipe, you must be explicit about the environment:

```{r}
env <- environment()
"x" %>% assign(100, envir = env)
x
```
    
Other functions with this problem include `get()` and `load()`.

2. Functions that use lazy evaluation.
In R, function arguments only computed if function uses them, not prior to calling the 
function. The pipe computes each element in turn, so you can't 
rely on this behaviour.
    
One place that this is a problem is `tryCatch()`, which lets you capture
and handle errors:
    
```{r, error = TRUE}
tryCatch(stop("!"), error = function(e) "An error")

stop("!") %>% 
  tryCatch(error = function(e) "An error")
```
    
There are a relatively wide class of functions with this behaviour,
including `try()`, `suppressMessages()`, and `suppressWarnings()`
in base R.
  

## When not to use the pipe

Pipes:

* Do not solve every problem
* are most useful for a fairly short linear sequence of operations

Think of another way to express yourself when:

* Many steps; create intermediate objects with meaningful names.
* Multiple inputs or outputs.
* Complex dependency structure: pipes are fundamentally linear and expressing 
  complex relationships with them will typically yield confusing code.

## Other tools from magrittr

* Side-effects: The "tee" pipe `%T>%`. 
  Like `%>%` except it returns the left-hand side instead of the right-hand side. 
  It's called "tee" because it's like a literal T-shaped pipe.

```{r}
rnorm(100) %>%
  matrix(ncol = 2) %>%
  plot() %>%
  str()

rnorm(100) %>%
  matrix(ncol = 2) %T>%
  plot() %>%
  str()
```

* `%$%`: "explodes" out the variables in a data frame so that you can 
    refer to them explicitly. 

```{r}
cor(mtcars$disp, mtcars$mpg)
mtcars %$% cor(disp, mpg)
with(mtcars, cor(disp, mpg))
```

