Guess what these functions do:

```
double f1(NumericVector x) {
   int n = x.size();
   double y = 0;
   
   for(int i = 0; i < n; ++i) {
     y += x[i];
   }
   
   return y / n;
}
```

```
NumericVector f2(NumericVector x) {
  int n = x.size();
  NumericVector out(n);
  out[0] = x[0];
  for(int i = 1; i < n; ++i) {
    out[i] = out[i - 1] + x[i];
  }
  return out;
}
```


```
bool f3(LogicalVector x) {
  int n = x.size();
  for(int i = 0; i < n; ++i) {
    if (x[i]) return true;
  }
  return false;
}
```





## Datatypes in Rcpp

See <http://adv-r.had.co.nz/Rcpp.html>

## Preparation

* `install.packages("Rcpp")`
* On windows, install [Rtools](https://cran.r-project.org/bin/windows/Rtools/)
* On mac: Xcode install command line tools.
* On linux, things work out of the box.

* Details can be found [here](http://www.rstudio.com/ide/docs/packages/prerequisites) 


To check your installation do:

```{r}
library(Rcpp)
evalCpp("1+1")
```

### Example: factorial

The factorial function 
$$n!=n\cdot(n-1)\cdot(n-2)\cdot\dots \cdot 2 \cdot 1$$ 
can be defined recursively as 
$$n!=f(n)=n f(n-1)$$ 
with $f(1)=1$.

Recursive implementation:

```{r}
fact.rec <- function(n){
  if (n == 1) return(1)
  return( n * fact.rec(n - 1) )
}
```


Iterative implementation:

```{r}
fact.it <- function(n){
  out <- 1
  if (n > 1){
      for (i in 2:n){
          out <- out * i
      }
  }
  out
}
```

Neither of the implementations are particularly fast.  
We benchmark the functions

```{r}
library(microbenchmark)
n <- 8

c(fact.rec(n), fact.it(n), factorial(n))

microbenchmark(fact.rec(n), fact.it(n), factorial(n))
```

Using Rcpp we can do:

```{r}
library(Rcpp)

# Recursive implementation
cppFunction(
'int fact_rec_c(int M){
     if (M==1) return 1;
     return M*fact_rec_c(M-1);
}')

# Iterative implementation
cppFunction('
int fact_it_c(int M){
    int out=1;
    if (M>1){
       for (int i=1; i<=M; ++i){
          out *= i;
       }
    }
    return out;
}')
```

```{r}
c( fact_rec_c(n), fact_it_c(n), factorial(n) )

microbenchmark(fact.rec(n), fact.it(n), 
               fact_rec_c(n), fact_it_c(n), factorial(n)  )
```


The `cppFunction()` above is excellent for testing things quickly, but soon we want our C++ code in a separate file. 
One way of creating this from within R is as:

```{r}
cat(
'#include <Rcpp.h>  
using namespace Rcpp ;

// [[Rcpp::export]]
int fact_rec_c(int M){
     if (M==1) return 1;
     return M*fact_rec_c(M-1);
}

// [[Rcpp::export]]
int fact_it_c(int M){
    int out=1;
    if (M>1){
       for (int i=1; i<=M; ++i){
          out *= i;
       }
    }
    return out;
}', file="factorial.cpp")
```

Compile the file with

```{r}
sourceCpp("factorial.cpp")
```

