---
title: "Data wrangling"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
```


# Data wrangling

Resources:

* <https://cran.rstudio.com/web/packages/dplyr/>
    * <https://cran.rstudio.com/web/packages/dplyr/vignettes/introduction.html>
* <http://r4ds.had.co.nz/transform.html>

Dataframes vs tibbles:
```{r, eval = FALSE}
mtcars
as_tibble(mtcars)
```


## The pipe operator, `%>%`

```{r}
filter(mpg, hwy >= 20, hwy <= 26)
```

```{r}
select(filter(mpg, hwy >= 20, hwy <= 26), manufacturer, hwy)
```

* Each data handling operation performed results in one more level of nested calls. 
* The code will quickly become difficult to read
    * Read it from the inner most call and out 
    * Operation description (function name) separated from arguments

The pipe operator. 

A crucial design decision of the `tidyverse`:

* First argument is a `tibble` (or `data.frame`)
* Subsequent arguments say what to do with the `tibble` (or `data.frame`)
* Always return a `tibble` (or `data.frame`)

This enables the so-called piping operator, `%>%`. Example:

# f(x, ...) === x %>% f(...)

```{r}
sum(sqrt(1:10))
1:10 %>% sqrt() %>% sum()
```

```{r}
filter(mpg, hwy >= 20, hwy <= 26) %>% 
  select(manufacturer, hwy)

mpg %>% 
  filter(hwy >= 20, hwy <= 26) %>% 
  select(manufacturer, hwy)
```

## Operations

* `filter()`: keep rows matching criteria
* `select()`: pick/reorder columns by name
* `rename()`: rename columns by name
* `arrange()`: reorder rows
* `mutate()`: add new or modify existing variables
* `summarise()`: perform aggregation operations
* `group_by()`: introduce aggregation groupings
* `count()`: a short-hand for `group_by()` and `summarise()`

A ton of helper functions.

## Examples

```{r}
mpg %>% select(manufacturer, hwy)
```

```{r}
mpg %>% select(class, everything())
```


```{r}
ggplot(mpg, aes(x = class)) + 
  geom_bar()

count(mpg, class)
mpg %>% count(class)
mpg %>% group_by(class) %>% summarise(n = n())

mpg %>% 
  group_by(class) %>% 
  summarise(n = n()) %>% 
  arrange(desc(n))

ggplot(mpg %>% count(class), aes(class, n)) +
  geom_col() # like geom_bar, but with height known
```


```{r}
mpg %>% 
  group_by(manufacturer) %>% 
  summarise(mean_hwy = mean(hwy)) %>% 
  arrange(desc(mean_hwy))

mpg %>% 
  group_by(cyl, trans) %>% 
  summarise(mean_hwy = mean(hwy)) %>% 
  arrange(cyl, desc(mean_hwy))
```

```{r}
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_point()
ggplot(mpg, aes(displ, hwy, color = class)) +
  geom_count()

mpg %>% 
  group_by(class, displ) %>% 
  summarise(mean_hwy = mean(hwy))

mpg %>% 
  group_by(class, displ) %>% 
  summarise(mean_hwy = mean(hwy)) %>% 
  ggplot(aes(displ, mean_hwy, color = class)) +
    geom_point()

d_tmp <- mpg %>% 
  group_by(class, displ) %>% 
  summarise(mean_hwy = mean(hwy))

ggplot(d_tmp, aes(displ, mean_hwy, color = class)) +
  geom_point()
```


## Line plots

```{r}
ggplot(d_tmp, aes(displ, mean_hwy, color = class)) +
  geom_point() +
  geom_line()
```



## Output

```{r}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_hwy = mean(hwy))
```

```{r}
mpg %>% 
  group_by(class) %>% 
  summarise(mean_hwy = mean(hwy)) %>% 
  pander::pander()
```

```{r}
library(pander)

mpg %>% 
  group_by(class) %>% 
  summarise(mean_hwy = mean(hwy)) %>% 
  pander(digits = 3)

mpg %>% 
  group_by(class) %>% 
  summarise(mean_hwy = mean(hwy)) %>% 
  pander(digits = 3)
```




# Databases

https://db.rstudio.com/dplyr/

```{r}
pass = ""
#pass = .rs.askForPassword("")
my_db <- src_postgres(host = 'localhost', 
                      dbname = 'nycflights13', 
                      user = 'tior', 
                      password = pass)
src_tbls(my_db)

tbl(my_db, sql("SELECT * FROM flights"))
tbl(my_db, sql("SELECT COUNT(*) FROM flights"))
tbl(my_db, sql("SELECT origin, COUNT(*) FROM flights GROUP BY origin"))

flights <- tbl(my_db, "flights")
flights

airports <- tbl(my_db, "airports")
airports

collect(airports)

f1 <- tbl(my_db, sql("SELECT * FROM flights"))
explain(f1)

f1 <- flights %>% 
  group_by(origin) %>% 
  summarise(n = n())
f1
explain(f1)
collect(f1)


d <- flights %>% 
  mutate(dep_hour = floor(dep_time/100)*100) %>% 
  count(dep_hour) %>% 
  arrange(dep_hour)
d
d %>% explain()
d %>% collect()

ggplot(d %>% collect(), aes(dep_hour, n)) +
  geom_col()
```
