---
title: | 
  | Miniproject in Introduction to Data Science
  | ITVEST Data Science and Big Data (DSBD)
output:
  pdf_document: 
    number_sections: true
    dev: png
  html_document: 
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, dpi = 300)
```

```{r, echo = TRUE}
library(tidyverse)
library(lubridate)
library(pander) # for prettier tables
library(scales) # for making prettier axes in plots
library(stringr)

theme_set(theme_bw())

panderOptions('big.mark', ',')
```

# Formalia

Deadline for hand-in: Jan 3, 2018 at 23:55.

Where: Moodle.

What: Rmd file. Possibly also pdf (or html), but Rmd is mandatory.

Groups: Maximum 3 participants, however the project must be handed in individually.

# Exercises

Here, we focus on the `airlines`, `airports`, `flights`, `planes`, and `weather`  datasets:

```{r, echo = TRUE}
library(nycflights13)

airlines
airports
flights
planes
weather
```

Remember to read about the datasets.

# Exercises


## Exercise

Construct a barplot displaying number of flights per month.

```{r}
flights %>% na.omit() %>% group_by(month) %>% count(month) %>% ggplot() + geom_bar(aes(month, n), stat="identity", color="blue") + labs(title="Flights per month", x="Month", y="Total flights") + scale_x_continuous(breaks = c(1:12))
```

Now, in the barplot (showing number of flights per month), make a separate bar for each origin.

```{r}
flights %>% na.omit() %>% group_by(month, origin) %>% summarize(count=n()) %>% ggplot() + geom_bar(aes(x=month, count, fill=origin), stat="identity", position="dodge") + labs(title="Flights per month", x="Month", y="Total flights") + scale_x_continuous(breaks=c(1:12))
```

## Exercise

What are the top-10 destinations and how many flights were made to these?

```{r}
flights %>% na.omit() %>% group_by(dest) %>% summarize(NoOfFlights=n()) %>% arrange(desc(NoOfFlights)) %>% top_n(10) %>% pandoc.table(big.mark=",", justify = c('left', 'right')) 
```

For these 10 destinations, make a barplot illustrating the number of flights from origin to top-10 destination.

```{r}
flights %>% na.omit() %>% group_by(dest) %>% summarize(NoOfFlights=n()) %>% arrange(desc(NoOfFlights)) %>% top_n(10) %>% 
  ggplot() + 
  geom_bar(aes(dest, NoOfFlights), stat="identity") + 
  labs(title = "Flights to top 10 destinations", x="Destinations", y="No. of flights")
```

Now, order the bars (destinations) according to the total number of flights to these destinations.

```{r}
flights %>% na.omit() %>% group_by(dest) %>% summarize(NoOfFlights=n()) %>% arrange(desc(NoOfFlights)) %>% top_n(10) %>% 
  ggplot() + 
  geom_bar(aes(factor(dest, levels = dest[order(-NoOfFlights)]), NoOfFlights), color="blue", stat="identity") + 
  labs(title = "Flights to top 10 destinations", x="Destinations", y="No. of flights")

```
## Exercise

Is the mean air time from each of the three origins different? Further, are these differences statistically significant?

```{r}
flightsMeanAirTimeByOrigin <- flights %>% na.omit() %>% group_by(origin) %>%  summarize(MeanAirTime =mean(air_time)) 
flightsMeanAirTimeByOrigin %>% pandoc.table(big.mark=",", justify = c('left', 'right'))


#TODO: Are the differences statistically diferent? use correlation to find out. i.e cor() function

```


## Exercise

How many weather observations are there for each origin?

```{r}
weather %>% na.omit() %>% group_by(origin) %>% summarize(Total=n()) %>% pandoc.table(big.mark=",", justify = c('left', 'right'))
```

Convert temperature to degrees Celsius. This will be used in the reminder of this miniproject.
(We do this for both `temp` and `dewp` using `mutate_at`)

```{r}
getWeatherTempInDeg <- function()
{
  weatherCleaned <- weather %>% filter(!is.na(temp)) %>%  mutate_at(vars(temp, dewp), funs(round((.-32)*5/9,  2)))
}
getWeatherTempInDeg()
```

Construct a graph displaying the temperature at `JFK`.

```{r}
jfkTemp <- getWeatherTempInDeg()  %>% filter(origin=='JFK') %>%  group_by(year, month, day) %>% summarize(deg=sum(temp))
plotJFKTemp <- function(){
  plot(jfkTemp$deg, col = "blue", main="Temperature at JFK", ylab="Daily Temp")
}
plotJFKTemp()
```

Add a red line showing the mean temperature for each day.

```{r}
plotJFKTemp()

#Calculate the daily mean temperature
jfkTemp <- getWeatherTempInDeg() %>% group_by(year, month, day) %>% filter(origin=='JFK') %>%  summarize(dailyMeanTemp = mean(temp))

abline(h=jfkTemp$dailyMeanTemp, col="red")

```

Now, visualize the daily mean temperature for each origin airport.

```{r}
#Get the daily mean temp
#Start by adding a date column of date type using lubridate's make_date function.
dailyMeanTempByOrigin <- getWeatherTempInDeg() %>% group_by(origin, day, month, year) %>% summarize(MeanTemp = mean(temp)) %>% mutate(date = make_date(year, month, day))

#Plot each day's graph against the mean temperature
dailyMeanTempByOrigin %>% na.omit() %>% ggplot() + geom_bar(aes(date, MeanTemp, fill=origin), stat="identity", position="dodge") + labs(title = "Daily mean temp. for each origin airport", x="Day", y="Mean temp. ºC") 
```

## Exercise

Investigate if arrival delay is associated with the flight distance (and also departure delay).

```{r}
#TODO: Make a reference to http://www.sthda.com/english/wiki/correlation-test-between-two-variables-in-r for better understanding

cor.test(~ arr_delay + distance, flights, method="pearson")
#The p-value is [< 2.2e-16] which is less than the significance level of 5%. This indicates that arr_delay and distance are significantly
# correlated with a coeff of -0.06 and p-value of 2.2^(-16). The relationship is however onot strong as indicated by a low coeff.

cor.test(~ dep_delay + distance, flights, method="pearson")
#The p-value is [< 2.2e-16] which is less than the significance level of 5%. This indicates that dep_delay and distance are significantly # correlated with a coeff of -0.02 and p-value of 2.2^(-16).  The relationship is however onot strong as indicated by a low coeff.

```
## Exercise

Investigate if departure delay is associated with weather conditions at the origin airport. This includes descriptives (mean departure delay), plotting, regression modelling, considering missing values etc.

```{r}

weatherCleaned <- weather %>% filter(!is.na(temp)) %>%  mutate_at(vars(temp, dewp), funs(round((.-32)*5/9,  2))) %>% group_by(year, month, day) %>% summarize(AvgTemp = mean(temp))

flightsCleaned <- flights %>% filter(!is.na(dep_delay)) %>% group_by(year, month, day) %>% summarize(MeanDepDelay=mean(dep_delay))
DelayComparedToWeather <- flightsCleaned %>% inner_join(weatherCleaned, c("year", "month", "day"))

#Plotting the departure delay against the temp.
plot(DelayComparedToWeather$AvgTemp, DelayComparedToWeather$MeanDepDelay, col="blue", main="Temp relation to dep. delay", xlab="Temp", ylab="Dep. delay")
lmfit <- lm(MeanDepDelay ~ AvgTemp, data = DelayComparedToWeather )
abline(lmfit)

#There is a fairly good relationship between the weather temp. and the departure delay.
```


## Exercise
Is the age of the plane associated to delay?

```{r}

flightsByYear <- flights %>% filter(!is.na(tailnum))
planesAgeDelayRelation <- planes %>% filter(!is.na(year)) %>% inner_join(flightsByYear, "tailnum") %>% filter(!is.na(dep_delay), !is.na(arr_delay)) %>% group_by(tailnum, year.x) %>% summarize(MeanDelay=mean(dep_delay)) %>% arrange(desc(MeanDelay))
k <- cor.test(planesAgeDelayRelation$MeanDelay, planesAgeDelayRelation$year.x, method="pearson")
plot(planesAgeDelayRelation$year.x, planesAgeDelayRelation$MeanDelay, col="blue", main="Delay compared to plane's age", xlab="Age", ylab="Delay in minutes")
abline(h=0.034)
#There is a weak relationship between the age of the plane and the delay as r is closer to 0


# 
# In the result below :
# 
# t is the t-test statistic value (t = 1.9419),
# df is the degrees of freedom (df= 3244),
# p-value is the significance level of the t-test (p-value = 0.5223^{-10}).
# conf.int is the confidence interval of the correlation coefficient at 95% (conf.int = [-0.0003286045, -0.0683986844]);
# sample estimates is the correlation coefficient (Cor.coeff = 0.03407532).
```

## Exercise

It seems like the plane manufacturer could use a cleaning. After that, how many manufacturers have more than 200 planes? And how many flights are each manufacturer with more than 200 planes responsible for?

```{r}
#Manufacturers with more than 200 planes -- After cleaning.
planesPostCleaning <- planes %>% filter(!is.na(manufacturer)) %>% group_by(manufacturer) %>% summarize(NoOfPlanes=n()) %>% filter(NoOfPlanes > 200) 

#Make pretty table
planesPostCleaning %>% pandoc.table(justify=c('left', 'right'))

#No of flights for manufacturers having more than 200 planes of which they are responsible thereof
planes %>% inner_join(flights, "tailnum") %>% filter(manufacturer %in% planesPostCleaning$manufacturer) %>% group_by(manufacturer) %>% summarize(NoOfFlights=n()) %>% pandoc.table(justify=c('left', 'right'))

```


## Exercise

It turns out that Airbus has several main models, e.g. several A320 (A320-211, A320-212 etc.) and so on. Create a frequency table of the main models for Airbus and how many planes there are in each.

```{r}
planes %>% filter(manufacturer=='AIRBUS') %>% group_by(model) %>% summarize(NoOfPlanes=n()) %>% arrange(NoOfPlanes) %>% pandoc.table(justify=c('left', 'right'))

```


## Exercise

Are larger planes (measured by number of seats) more or less delayed than smaller planes?

```{r}
flightsDelay <- flights %>% filter(!is.na(arr_delay), !is.na(dep_delay)) %>% mutate(totalDelay=arr_delay+dep_delay)
planesDelayBySize <- planes %>% filter(!is.na(manufacturer), !is.na(seats)) %>%  inner_join(flightsDelay, "tailnum")
planesDelayBySize <- planesDelayBySize %>% group_by(seats) %>% summarize(AvgDelay=mean(totalDelay)) 


#Inspection
scatter.smooth(planesDelayBySize$seats, planesDelayBySize$AvgDelay, col="red", main="Planes delay by no. of seats", xlab="No. of setas", ylab="Avg. delay (min)")

#There is no linear relationship between the size of plane (No. of seats) and the delay

```


## Exercise

On a map (`map_data("usa")`), plot the airports that has flights to them.

```{r}
usa <- map_data("usa") 
airportsWithFlights <- airports %>% na.omit() %>% inner_join(flights, c("faa"= "dest")) %>% group_by(faa, lat, lon) %>% summarize(NoOfFlights=n())

ggplot() + 
  geom_polygon(data = usa, aes(long, lat, group = group), fill = "grey") + 
  geom_point(data = airportsWithFlights , aes(lon, lat), color="red") + 
  labs(title="Airports that have flights") + 
  coord_quickmap()
#101 airports have flights to them. I.e they have dest.
```

Make a similar plot, but now points must have size relative to the number of flights each airport is destination for. 

```{r}
ggplot() + 
  geom_polygon(data = usa, aes(long, lat, group = group), fill = "grey") + 
  geom_point(data = airportsWithFlights , aes(lon, lat, size=NoOfFlights), position="jitter") + 
  labs(title="Airports that have flights") + 
  coord_quickmap()
```

## Exercise

Do a principal component analysis of the weather at JFK using the following columns: `temp, dewp, humid, wind_dir, wind_speed, precip, visib` (only on `complete.cases()`). How many principal components should be used to capture the variability in the weather data? 

```{r}

```

